---
title: "Advanced Spatial Visualization: Minnesota Mapping"
subtitle: "ICA 04 — Spatial Data with `sf`, Elevation Rasters, and Multi-Layer Maps"
author: "Zhijun He"
date: "2026-02-11"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Data Story Critique

**Source:** [USGS Drought Timeline Visualization](https://labs.waterdata.usgs.gov/visualizations/drought-timeline/index.html#/)

**What is the data story?** The visualization tracks the spatial and temporal evolution of drought conditions across the United States, showing how drought intensity and coverage have changed over time at the county level.

**What is effective?** The timeline scrubber allows users to animate drought progression, making temporal patterns intuitive. The color encoding for drought severity (D0–D4) follows a clear sequential palette, and the geographic granularity at the county level reveals regional patterns that state-level data would miss.

**What could be improved?** The visualization could benefit from an accompanying line chart showing aggregate drought area over time, giving users a macro-level summary alongside the geographic detail. Adding the ability to click on a county for a local time series would also enhance exploratory analysis.

## Setup

```{r load-packages}
# Load required packages
library(tidyverse)
library(sf)
library(elevatr)
library(terra)
library(stars)
library(tidycensus)
library(USAboundaries)
library(ggthemes)
```

## Reading Spatial Data

```{r read-data}
# Built-in North Carolina shapefile
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

# MN cities (population centers)
mn_cities <- st_read("/Users/moonhalo/Documents/GitHub/portfolio-Zhijun1933-212/src/ica/04_adv_maps/data/shp_loc_pop_centers/city_and_township_population_centers.shp", quiet = TRUE)

# MN water (lakes and rivers)
mn_water <- st_read("/Users/moonhalo/Documents/GitHub/portfolio-Zhijun1933-212/src/ica/04_adv_maps/data/shp_water_lakes_rivers/LakesAndRivers.shp", quiet = TRUE)
```

### Checking Object Classes

The `sf` package reads spatial data into a hybrid `sf` / `data.frame` object:

```{r check-classes}
class(nc)
class(mn_cities)
class(mn_water)
```

### Checking CRS

```{r check-crs}
st_crs(nc)
```

### Plotting NC in Original CRS (NAD27)

```{r nc-plot-original}
ggplot(nc) +
  geom_sf() +
  theme_classic() +
  labs(title = "North Carolina Counties in NAD27 CRS")
```

## Changing the CRS

Transforming to a South Carolina CRS to observe distortion:

```{r nc-transform-sc}
nc_transformed <- nc |> st_transform(crs = "EPSG:32133")
st_crs(nc_transformed)

ggplot(nc_transformed) +
  geom_sf() +
  theme_classic() +
  labs(title = "North Carolina in EPSG:32133 (South Carolina State Plane)")
```

### Custom Function with Title Argument

```{r transform-function}
transform_and_plot <- function(spatial_obj, new_crs, title = "Transformed Map") {
  spatial_obj |>
    st_transform(crs = new_crs) |>
    ggplot() +
    geom_sf() +
    theme_classic() +
    labs(title = title)
}
```

#### CRS 1: Alaska Albers (EPSG:3338) — Large-Scale Distortion

```{r crs1}
transform_and_plot(nc, new_crs = "EPSG:3338",
                   title = "NC Projected in Alaska Albers (EPSG:3338)")
```

#### CRS 2: Web Mercator (EPSG:3857) — Common Web Map Projection

```{r crs2}
transform_and_plot(nc, new_crs = "EPSG:3857",
                   title = "NC Projected in Web Mercator (EPSG:3857)")
```

**Verify your understanding:** If point location data were stored in a CRS different from NAD27, overlaying them directly on the NAD27 map would produce spatial misalignment — points would appear shifted from their true locations because the coordinate systems define position differently. We must first transform all layers to a common CRS using `st_transform()`.

## Minnesota Map with Multiple Layers

```{r install-usboundaries, eval=FALSE}
install.packages("USAboundaries", repos = "https://ropensci.r-universe.dev", type = "source")
install.packages("USAboundariesData", repos = "https://ropensci.r-universe.dev", type = "source")
```
```{r mn-counties}
mn_counties <- USAboundaries::us_counties(resolution = "high", states = "Minnesota")

# Fix duplicate column names
names_counties <- names(mn_counties)
names(mn_counties)[names_counties == "state_name"] <- c("state_name1", "state_name2")
```

### Unifying CRSs

```{r check-crs-all}
cat("mn_cities CRS:\n")
st_crs(mn_cities)$input
cat("\nmn_water CRS:\n")
st_crs(mn_water)$input
cat("\nmn_counties CRS:\n")
st_crs(mn_counties)$input
```

```{r unify-crs}
# Transform all datasets to match mn_cities CRS
mn_counties <- mn_counties |> st_transform(crs = st_crs(mn_cities))
mn_water <- mn_water |> st_transform(crs = st_crs(mn_cities))
```

### Counties + Cities

```{r counties-cities}
ggplot() +
  geom_sf(data = mn_counties, fill = "white", color = "gray40") +
  geom_sf(data = mn_cities, size = 0.3, color = "steelblue", alpha = 0.5) +
  labs(title = "Minnesota City/Township Population Centers by County") +
  ggthemes::theme_map()
```

### Customizing Colors: Large Cities by Population

```{r cities-population}
ggplot() +
  geom_sf(data = mn_counties, fill = "white") +
  geom_sf(
    data = mn_cities |> filter(Population >= 10000),
    mapping = aes(color = Population, size = Population)
  ) +
  scale_color_viridis_c() +
  labs(title = "Minnesota Cities with Population ≥ 10,000") +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
```

**`scale_color_viridis_c()` notes:**

The viridis scales are perceptually uniform, meaning they are readable in color, in greyscale, and by viewers with common forms of color blindness. The `_d()` variant handles discrete/categorical variables, `_c()` handles continuous variables, and `_b()` bins a continuous variable into a small number of discrete color steps.

### Adding Elevation Raster Data

```{r elevation-mn, results='hide'}
elevation <- elevatr::get_elev_raster(mn_counties, z = 5, clip = "bbox")
raster::crs(elevation) <- sf::st_crs(mn_counties)

elev_df <- elevation |> terra::as.data.frame(xy = TRUE)
colnames(elev_df) <- c("x", "y", "elevation")
```

```{r elevation-map}
ggplot() +
  geom_raster(data = elev_df, aes(x = x, y = y, fill = elevation)) +
  scale_fill_gradient(low = "darkgreen", high = "white", guide = "none") +
  geom_sf(data = mn_counties, fill = NA, color = "black") +
  geom_sf(
    data = mn_cities |> filter(Population >= 10000),
    mapping = aes(color = Population, size = Population)
  ) +
  scale_color_viridis_c() +
  labs(title = "Minnesota Elevation with Cities ≥ 10,000 Population") +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
```

### Zoom in to Twin Cities and Add Water

```{r twin-cities-bbox}
seven_countyarea <- mn_counties |>
  filter(name %in% c("Anoka", "Hennepin", "Ramsey", "Dakota", "Carver", "Washington", "Scott")) |>
  st_bbox()
seven_countyarea
```

```{r twin-cities-elevation, results='hide'}
elevation <- elevatr::get_elev_raster(
  mn_counties |> st_crop(seven_countyarea), z = 9, clip = "bbox"
)
raster::crs(elevation) <- sf::st_crs(mn_counties)

elev_df <- elevation |> terra::as.data.frame(xy = TRUE)
colnames(elev_df) <- c("x", "y", "elevation")
```

```{r twin-cities-map}
tc_map <- ggplot() +
  geom_raster(data = elev_df, aes(x = x, y = y, fill = elevation)) +
  geom_sf(data = mn_counties, fill = NA, color = "black") +
  geom_sf(data = mn_water, fill = "lightsteelblue1", color = "lightsteelblue1") +
  geom_sf(
    data = mn_cities |> filter(Population >= 10000),
    mapping = aes(color = Population, size = Population)
  ) +
  scale_color_viridis_c(option = "magma") +
  scale_fill_gradient(low = "darkgreen", high = "white") +
  coord_sf(
    xlim = seven_countyarea[c("xmin", "xmax")],
    ylim = seven_countyarea[c("ymin", "ymax")]
  ) +
  labs(title = "Twin Cities Metro: Elevation, Water, and Cities ≥ 10,000 Population") +
  ggthemes::theme_map() +
  theme(legend.position = "none")

tc_map
```

### Saving the Map

```{r save-map}
png("/Users/moonhalo/Documents/GitHub/portfolio-Zhijun1933-212/src/ica/04_adv_maps/figures/tc_map_zoom.png", width = 800, height = 400)
tc_map
dev.off()
```

## Going Beyond: Twin Cities with Leaflet

```{r leaflet-map}
library(leaflet)

mn_counties_leaf <- mn_counties |> st_transform(4326)
mn_cities_leaf <- mn_cities |> st_transform(4326)

cities_per_county <- st_join(mn_cities_leaf, mn_counties_leaf) |>
  st_drop_geometry() |>
  count(name)

mn_counties_leaf |>
  filter(name %in% c("Anoka", "Hennepin", "Ramsey", "Dakota", "Carver", "Washington", "Scott")) |>
  left_join(cities_per_county) |>
  leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    color = "#444444", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5,
    fillColor = ~colorQuantile("YlOrRd", n)(n),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE)
  ) |>
  addCircles(
    data = mn_cities_leaf |>
      filter(County %in% paste(
        c("Anoka", "Hennepin", "Ramsey", "Dakota", "Carver", "Washington", "Scott"), "County"
      )),
    color = "#444444"
  )
```

---

*Zhijun He | Macalester College | `r Sys.Date()`*
