---
title: "The Paradox of Brazil's Micro-Enterprises"
subtitle: "How the smallest firms hold the least capital despite dominating the corporate landscape"
author: "Zhijun He"
date: today
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

## Research Question

**Do micro-enterprises in Brazil face a "capital trap"—where they constitute the majority of registered firms yet control a disproportionately small share of total capital stock?**

This question matters for development economics: if the smallest formal businesses cannot accumulate capital at rates proportional to their numbers, it suggests structural barriers to firm growth that may perpetuate economic inequality and limit broad-based development.

## Background

Brazil's CNPJ registry provides a rare window into the formal business sector of Latin America's largest economy. The dataset contains companies filtered to those with capital stock exceeding 150,000 BRL (~$30,000 USD), meaning we're examining firms that have already crossed a meaningful capitalization threshold. If capital concentration exists even within this subset of more substantial businesses, it points to deeper structural patterns in how Brazilian firms accumulate resources.

## Data & Analysis

```{r}
#| label: load-packages

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(scales)
library(stringr)
library(showtext)

# Load Google Font for professional typography
font_add_google("Source Sans Pro", "source")
showtext_auto()
```

```{r}
#| label: load-data

# Load data (comma-delimited)
companies <- read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv',
  show_col_types = FALSE
)

# Prepare summary data
size_summary <- companies |>
  dplyr::filter(!is.na(company_size), !is.na(capital_stock)) |>
  dplyr::group_by(company_size) |>
  dplyr::summarise(
    n_companies = n(),
    total_capital = sum(capital_stock, na.rm = TRUE),
    mean_capital = mean(capital_stock, na.rm = TRUE),
    median_capital = median(capital_stock, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pct_companies = n_companies / sum(n_companies) * 100,
    pct_capital = total_capital / sum(total_capital) * 100,
    capital_gap = pct_companies - pct_capital,
    company_size = dplyr::case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE ~ "Medium/Large"
    ),
    company_size = factor(company_size, levels = c("Micro", "Small", "Medium/Large"))
  )
```

```{r}
#| label: fig-capital-paradox
#| fig-cap: "Brazil's Capital Paradox: Micro-enterprises make up the majority of firms but control a fraction of total capital"
#| fig-width: 10
#| fig-height: 7

# Reshape for visualization
plot_data <- size_summary |>
  dplyr::select(company_size, pct_companies, pct_capital) |>
  tidyr::pivot_longer(
    cols = c(pct_companies, pct_capital),
    names_to = "metric",
    values_to = "percentage"
  ) |>
  dplyr::mutate(
    metric = dplyr::if_else(metric == "pct_companies", "Share of Companies", "Share of Capital"),
    metric = factor(metric, levels = c("Share of Companies", "Share of Capital"))
  )

# Define colors
colors <- c("Share of Companies" = "#264653", "Share of Capital" = "#e76f51")

# Create the visualization
ggplot(plot_data, aes(x = company_size, y = percentage, fill = metric)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
  geom_text(
    aes(label = sprintf("%.1f%%", percentage)),
    position = position_dodge(width = 0.75),
    vjust = -0.5,
    size = 4.5,
    fontface = "bold",
    family = "source"
  ) +
  # Add annotation for the key insight
  annotate(
    "segment",
    x = 0.65, xend = 0.65,
    y = plot_data |> dplyr::filter(company_size == "Micro", metric == "Share of Capital") |> dplyr::pull(percentage),
    yend = plot_data |> dplyr::filter(company_size == "Micro", metric == "Share of Companies") |> dplyr::pull(percentage),
    color = "#2a9d8f",
    linewidth = 1.5,
    arrow = arrow(ends = "both", length = unit(0.15, "inches"))
  ) +
  annotate(
    "text",
    x = 0.45,
    y = (plot_data |> dplyr::filter(company_size == "Micro", metric == "Share of Capital") |> dplyr::pull(percentage) +
         plot_data |> dplyr::filter(company_size == "Micro", metric == "Share of Companies") |> dplyr::pull(percentage)) / 2,
    label = "Capital\nGap",
    hjust = 1,
    size = 3.5,
    color = "#2a9d8f",
    fontface = "bold",
    family = "source"
  ) +
  scale_fill_manual(values = colors, name = NULL) +
  scale_y_continuous(
    limits = c(0, 85),
    breaks = seq(0, 80, 20),
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Brazil's Micro-Enterprise Capital Trap",
    subtitle = "Micro-enterprises dominate firm counts but control far less capital,\nwhile medium/large firms show the opposite pattern",
    x = NULL,
    y = NULL,
    caption = "Source: Brazilian CNPJ Registry (Receita Federal) via TidyTuesday 2026 Week 4\nNote: Dataset filtered to companies with capital stock > 150,000 BRL"
  ) +
  theme_minimal(base_size = 14, base_family = "source") +
  theme(
    plot.title = element_text(face = "bold", size = 20, color = "#264653", margin = margin(b = 5)),
    plot.subtitle = element_text(size = 13, color = "gray40", margin = margin(b = 20), lineheight = 1.2),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0, margin = margin(t = 15)),
    legend.position = "top",
    legend.justification = "left",
    legend.text = element_text(size = 12),
    legend.margin = margin(b = 10),
    axis.text = element_text(size = 12, color = "gray30"),
    axis.text.x = element_text(face = "bold", size = 13),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray90"),
    plot.margin = margin(20, 20, 20, 20)
  )
```

## Visualization 2: Capital Inequality Landscape

This advanced visualization combines multiple dimensions: company size, legal nature, and the Gini-like concentration of capital within each category.

```{r}
#| label: fig-advanced-viz
#| fig-cap: "Capital concentration across Brazil's corporate landscape: bubble size shows total capital, position reveals the gap between firm count and capital share"
#| fig-width: 12
#| fig-height: 9

# Calculate metrics by company size AND legal nature (top categories)
top_legal <- companies |>
  dplyr::count(legal_nature, sort = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::pull(legal_nature)

combined_summary <- companies |>
  dplyr::filter(
    !is.na(company_size), 
    !is.na(capital_stock),
    !is.na(legal_nature),
    legal_nature %in% top_legal
  ) |>
  dplyr::group_by(company_size, legal_nature) |>
  dplyr::summarise(
    n_companies = dplyr::n(),
    total_capital = sum(capital_stock, na.rm = TRUE),
    mean_capital = mean(capital_stock, na.rm = TRUE),
    median_capital = median(capital_stock, na.rm = TRUE),
    # Capital concentration ratio (mean/median - higher = more skewed)
    concentration_ratio = mean_capital / median_capital,
    .groups = "drop"
  ) |>
  # Calculate percentages within each legal nature
  dplyr::group_by(legal_nature) |>
  dplyr::mutate(
    pct_companies = n_companies / sum(n_companies) * 100,
    pct_capital = total_capital / sum(total_capital) * 100,
    capital_gap = pct_capital - pct_companies  # Positive = capital advantage
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    company_size = dplyr::case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE ~ "Medium/Large"
    ),
    company_size = factor(company_size, levels = c("Micro", "Small", "Medium/Large")),
    # Shorten legal nature names
    legal_nature_short = dplyr::case_when(
      stringr::str_detect(legal_nature, "LLC") ~ "LLC",
      stringr::str_detect(legal_nature, "Individual") ~ "Individual\nEntrepreneur",
      stringr::str_detect(legal_nature, "Corporation") ~ "Corporation",
      stringr::str_detect(legal_nature, "Simple") ~ "Simple\nCompany",
      TRUE ~ stringr::str_trunc(legal_nature, 15)
    )
  )

# Create the advanced visualization
ggplot(combined_summary, 
       aes(x = pct_companies, y = pct_capital)) +
  # Add diagonal reference line (equal distribution)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray60", linewidth = 0.8) +
  # Add segments from point to diagonal (showing the gap)
  geom_segment(
    aes(xend = pct_companies, yend = pct_companies, color = company_size),
    linewidth = 0.8, alpha = 0.5
  ) +
  # Add bubbles

  geom_point(
    aes(size = total_capital / 1e9, fill = company_size),
    shape = 21, alpha = 0.85, stroke = 1.2, color = "white"
  ) +
  # Add text labels for concentration ratio
  geom_text(
    aes(label = sprintf("%.1fx", concentration_ratio)),
    size = 2.8, fontface = "bold", color = "gray20",
    vjust = -1.8, family = "source"
  ) +
  # Facet by legal nature

  facet_wrap(~legal_nature_short, nrow = 1) +
  # Scales
  scale_x_continuous(
    limits = c(0, 100),
    breaks = c(0, 50, 100),
    labels = function(x) paste0(x, "%")
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = c(0, 50, 100),
    labels = function(x) paste0(x, "%")
  ) +
  scale_size_continuous(
    range = c(4, 25),
    labels = scales::label_number(suffix = "B BRL"),
    name = "Total Capital",
    breaks = c(1, 10, 50, 100)
  ) +
  scale_fill_manual(
    values = c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653"),
    name = "Company Size"
  ) +
  scale_color_manual(
    values = c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653"),
    guide = "none"
  ) +
  # Labels
  labs(
    title = "Where Capital Concentrates: A Multi-Dimensional View",
    subtitle = "Points above the diagonal control more capital than their numbers suggest | Labels show mean/median ratio (higher = more internal inequality)",
    x = "Share of Companies →",
    y = "← Share of Capital",
    caption = "Source: Brazilian CNPJ Registry via TidyTuesday 2026 Week 4 | Bubble size = total capital stock"
  ) +
  # Annotations
  annotate(
    "text", x = 85, y = 15, 
    label = "Capital\nDeficit\nZone", 
    color = "#e76f51", size = 3.5, fontface = "italic",
    family = "source", lineheight = 0.9
  ) +
  annotate(
    "text", x = 15, y = 85, 
    label = "Capital\nSurplus\nZone", 
    color = "#264653", size = 3.5, fontface = "italic",
    family = "source", lineheight = 0.9
  ) +
  # Theme
  theme_minimal(base_size = 13, base_family = "source") +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#264653", margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "gray40", margin = margin(b = 15), lineheight = 1.2),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0, margin = margin(t = 15)),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.margin = margin(t = 10),
    strip.text = element_text(face = "bold", size = 12, color = "#264653"),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.spacing = unit(1.5, "lines"),
    axis.title = element_text(size = 11, color = "gray40"),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  guides(
    fill = guide_legend(order = 1, override.aes = list(size = 6)),
    size = guide_legend(order = 2)
  )
```

### Reading This Visualization

This chart reveals capital inequality through multiple lenses:

- **Position**: Points above the diagonal line control *more* capital than their share of companies would suggest (capital surplus); points below control *less* (capital deficit)
- **Vertical lines**: Show the magnitude of the gap between expected and actual capital share
- **Bubble size**: Represents absolute total capital in billions of BRL
- **Labels (e.g., "2.5x")**: The mean/median ratio—a proxy for internal inequality. Higher values mean a few firms within that category hold most of the capital
- **Facets**: Compare patterns across Brazil's main legal structures

## Key Finding

The visualization reveals a stark **capital paradox** in Brazil's corporate structure:

- **Micro-enterprises** constitute the largest share of registered companies yet control a disproportionately small fraction of total capital stock
- **Medium/Large firms** show the inverse pattern—fewer in number but commanding the majority of capital
- The "capital gap" (the difference between company share and capital share) is most pronounced for micro-enterprises

This pattern persists even after filtering for firms with capital above 150,000 BRL, suggesting that the barriers to capital accumulation for small firms are structural rather than simply a function of including marginal businesses.

## Implications

For development practitioners and policymakers, this finding raises important questions:

1. **Access to finance**: Do micro-enterprises face credit constraints that prevent capital accumulation?
2. **Regulatory burden**: Are compliance costs disproportionately affecting smaller firms' ability to reinvest?
3. **Market structure**: Does Brazil's economic structure favor consolidation over broad-based firm growth?

The concentration of capital in larger firms—even within a dataset of already-formalized, reasonably capitalized businesses—suggests that Brazil's development challenge isn't just about formalizing the informal sector, but about enabling smaller formal firms to grow.
